---
title: "summary_table"
output: pdf_document
date: "2024-08-29"
---
```{r}
# Load the datasets
meta_data_path = 'dataset/QBS103_GSE157103_series_matrix.csv'
gene_data_path = 'dataset/QBS103_GSE157103_genes.csv'
genes_data <- read_csv(gene_data_path, show_col_types = FALSE)
metadata <- read_csv(meta_data_path, show_col_types = FALSE)
colnames(metadata)
```

```{r}
metadata <- metadata %>%
  mutate(
    age = as.numeric(`age`),
    ferritin = as.numeric(`ferritin(ng/ml)`),
    crp = as.numeric(`crp(mg/l)`),
  )
```

```{r}
summary_continuous <- function(var, group) {
  metadata %>%
    group_by(!!sym(group)) %>%
    summarise(
      Mean = mean(!!sym(var), na.rm = TRUE),
      SD = sd(!!sym(var), na.rm = TRUE),
      Median = median(!!sym(var), na.rm = TRUE),
      IQR = IQR(!!sym(var), na.rm = TRUE)
    )
}

age_summary <- summary_continuous("age", "icu_status")
ferritin_summary <- summary_continuous("ferritin", "icu_status")
crp_summary <- summary_continuous("crp", "icu_status")
age_summary
ferritin_summary
crp_summary
```

```{r}
summary_categorical <- function(var, group) {
  metadata %>%
    group_by(!!sym(group), !!sym(var)) %>%
    summarise(n = n()) %>%
    mutate(percentage = n / sum(n) * 100)
}
sex_summary <- summary_categorical("sex", "icu_status")
charlson_summary <- summary_categorical("disease_status", "icu_status")
sex_summary
charlson_summary
```

```{r}

```{r, warning=FALSE, message=FALSE}
# Define a custom theme
custom_theme <- theme(
    panel.background = element_rect(fill = '#333333'),
    plot.title = element_text(hjust = 0.5, color = 'white', size = 16),
    axis.title = element_text(size = 14, color = 'white'),
    axis.text = element_text(size = 12, color = 'white'),
    plot.background = element_rect(fill = '#333333'),
    panel.grid.major = element_line(color = "gray", size = 0.1),
    legend.position = 'top',
    legend.background = element_rect(fill = "#333333"),
    legend.text = element_text(size = 12, color = 'white'),
    legend.title = element_text(size = 12, color = 'white'),
    legend.key = element_rect(fill = "#333333")
)
```

```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
# Define the function to create plots
create_plots <- function(meta_data_path, gene_data_path, gene_names, 
                         continuous_covariate, categorical_covariate1, 
                         categorical_covariate2) {
    # Load the datasets
    genes_data <- read_csv(gene_data_path, show_col_types = FALSE)
    metadata <- read_csv(meta_data_path, show_col_types = FALSE)
    
    # Check if the data is read properly
    if (is.null(genes_data) || is.null(metadata)) {
      stop("Failed to read the datasets")
    }
    
    # Rename the column `...1` to `Genes` if it exists
    if ("...1" %in% colnames(genes_data)) {
        genes_data <- genes_data %>% rename(Genes = `...1`)
    } else {
        stop("Column '...1' not found in gene data")
    }

    # Set gene names as row names
    rownames(genes_data) <- genes_data$Genes

    for (gene_name in gene_names) {
        gene_name <- as.character(gene_name)
        #print(paste("Processing gene:", gene_name))
        if (!(gene_name %in% rownames(genes_data))) {
            warning(paste("Gene name", gene_name, "not found in the genes data"))
            next
        }

      # Select the gene row and transpose
      gene <- genes_data[gene_name, ]
      transposed_gene <- as.data.frame(t(gene))
      colnames(transposed_gene) <- gene_name
      # Combine two datasets with participant_id and gene
      transposed_gene_data <- cbind(participant_id = rownames(transposed_gene), transposed_gene)
      # Merge the dataset
      data_frame <- merge(metadata, transposed_gene_data, by = "participant_id")
      # Ensure the gene expression values are numeric
      data_frame[[gene_name]] <- as.numeric(data_frame[[gene_name]])
      # Filter out NA values for the continuous covariate
      data_frame[[continuous_covariate]] <- as.numeric(data_frame[[continuous_covariate]])
      data_frame <- data_frame %>% filter(!is.na(!!sym(continuous_covariate))) 
      # Filter out NA values for categorical covariate
      data_frame <- data_frame %>% filter((!!sym(categorical_covariate1) != 'unknown'))
      filtered_data <- data_frame %>% filter((!!sym(categorical_covariate2) != 'unknown'))
      # Histogram for gene expression
      gene_histogram <- ggplot(filtered_data, aes(x = !!sym(gene_name))) + 
          geom_histogram(binwidth = 0.05, color = "#1dddeb", fill = "#1dddeb", alpha = 0.7) + 
          labs(x = gene_name, 
               y = "Frequency", 
               title = paste("Histogram of",
                             gene_name, 
                             "Expression")) +
          custom_theme
      
      # Save the histogram
      ggsave(paste0("plots/", gene_name, "_histogram.png"), 
             plot = gene_histogram, width = 16, height = 9)
      # Print the histogram
      print(gene_histogram)
      
      # Scatter Plot with rainbow colors
      scatter_plot <- ggplot(filtered_data, 
                             aes(x = !!sym(continuous_covariate),
                                 y = !!sym(gene_name), 
                                 color = !!sym(gene_name))) + 
          geom_point(size = 2) + 
          geom_rug(color = "white") +  # Add a rug plot
          scale_color_gradientn(colors = rainbow(7)) +  # Apply rainbow colors
          labs(x = continuous_covariate, y = gene_name, title = paste("Scatter Plot of",
                                                                      gene_name, 
                                                                      "Expression vs", 
                                                                      continuous_covariate)) +
          custom_theme
      
      # Save the scatter plot
      ggsave(paste0("plots/", gene_name, "_scatter_plot.png"), 
             plot = scatter_plot, width = 16, height = 9)
      # Print the scatter plot
      print(scatter_plot)
      # Define labels for the legend
      labels <- unique(filtered_data[[categorical_covariate2]])
      # Create the palette dynamically based on the labels
      color <- c("#66c2a5", "#fc8d62")
      palette <- setNames(color, labels)
      # https://www.statology.org/setnames-r/
      # Boxplot
      boxplot <- ggplot(filtered_data, aes(x = !!sym(categorical_covariate1), 
                                           y = !!sym(gene_name), 
                                           fill = !!sym(categorical_covariate2))) + 
          geom_boxplot(outlier.color = "red", color = "white") + 
          labs(x = categorical_covariate1, y = gene_name, fill = categorical_covariate2, 
               title = paste("Boxplot of", gene_name, "Expression by", 
                             categorical_covariate1, "and", categorical_covariate2)) +
          scale_fill_manual(values = palette, labels = labels) + 
          custom_theme
      
      # Save the boxplot
      ggsave(paste0("plots/", gene_name, "_boxplot.png"), 
             plot = boxplot, width = 16, height = 9)
      # Print the boxplot
      print(boxplot)
    }
}

create_plots('dataset/QBS103_GSE157103_series_matrix.csv',
             'dataset/QBS103_GSE157103_genes.csv',
             c("A1BG"), 
             "age", "sex", 
             "disease_status")
```

```